The Correspondence Analysis of Two Mode Networks Revisited

### 2025-03-25 main-R1.tex
Updates from Overleaf

\begin{figure}[ht!]
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/ca-reord.png}
            \caption{}
            \label{fig:ca-reord}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord1.png}
            \caption{}
            \label{fig:bon-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord2.png}
            \caption{}
            \label{fig:bon-reord2}
    \end{subfigure}
    \caption{Southern Women Affiliation matrix with re-ordered rows and columns: (a) row and columns re-ordered according to the person and group score on the first CA dimension, (b) row and columns re-ordered according to person and group Eigenvector centrality score, (c) row and columns re-ordered according to person and group score on the second Bonacich dimension.}
    \label{fig:ca-v-bon}
\end{figure}

@@

The larger point is that the standard eigenvector and CA scores extract qualitatively different information from the two-mode network, with $C^R$ geared toward a similarity-based partitioning---with the exact kind of similarity revealed by CA yet to be determined---and $C^B$ more focused on finding a core of well-connected people and groups, separated from a less well-connected periphery.

@@

This suggests that the eigenvector centralities do encode a kind of similarity between pairs of people and pairs of groups; it just happens to be the unweighted similarity scores given by their raw number of shared memberships (for people) and shared members (for groups). This also means that the dual centrality eigenvector score mixes an ordinal rank criterion with a similarity criterion since people with lots of members will end up being similar to lots of other people (and the same for large groups). This also means that higher order axes of the Bonacich centrality eigendecomposition of the Breiger projection matrices should also encode other forms of similarity between pairs of people and pairs of groups, independently of the main axis dominated by overall person activity and group size.

While plotting the Bonacich centrality against higher eigenvectors is not necessarily a substantively unmotivated practice---\citep[see][for a discussion in the one-mode network case]{iacobucci2017eigenvector}---the exact meaning of this mysterious second eigenvector remains unclear, especially in the two-mode case. We can use the same ``a posteriori blockmodeling'' \citep{wasserman1990correspondence} approach we used before to contrast CA and the Bonacich centrality analysis to shed light on this issue. Figure~\ref{fig:bon-reord2} shows the affiliation matrix blocked according to the \textit{second} Bonacich dimension, and Figure~\ref{fig:bon-sim} (c) shows the ``Bonacich Correspondence Plot'' with the usual eigenvector centrality score on the x-axis plotted against the second eigenvector of the Breiger projection matrix for both persons and groups on the y-axis.

We can see from Figure~\ref{fig:bon-reord2} that the second Bonacich dimension reveals a \textit{dual triangular} structure, revealing two secondary core and peripheries that are not encoded in either the first Bonacich or CA dimensions. Interestingly, this partition is almost equivalent to that obtained in \citet[Table 5]{everett2013dual} by clustering of a version of the projection matrices where the proximities were designed to detect structural equivalence (e.g., the cell entry in each projection matrix was the proportion of matches between each pair of people and groups).\footnote{As \citet[208]{everett2013dual} define it, ``[p]roportion matching is the number of times two profiles have the same entry (either zeros or ones) divided by the length of the profile.''} In Everett and Borgatti's \citeyearpar[208]{everett2013dual} clustering solution, ``[t]he women are then split into (a) those that
attend four or fewer events and consequently attend very few of the peripheral events (Pearl, Ruth, Verne, Myrna [sic], Dorothy, Olivia, Flora), and (b) those that attend core events and the first group of peripheral events (Evelyn, Laura, Theresa, Brenda, Charlotte, Francis, Eleanor), and (c) those that attend core events and the second group of peripheral events (Helen, Katherine, Sylvia, Nora).'' Note that the only disagreement between Everett and Borgatti's partition and the three-block solution suggested by Figure~\ref{fig:bon-reord2} is the position of Myrna/Myra's, who is assigned to Borgatti and Everett's third group in the a posteriori blocking obtained from Bonacich's second dimension.

This analysis points to a new way to interpret the second Bonacich dimension, namely, as a way to reveal secondary core-periphery partitions net of a main core-periphery partition in the two-mode network (if any exists). Thus, in the Bonacich Correspondence Plot shown in Figure~\ref{fig:bon-sim}(c), we can see that persons and groups are ordered by their ``main'' core-periphery scores from left-to-right, while the vertical segregation reveals

>>>>

\begin{figure}[ht!]
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/ca-reord1.png}
            \caption{}
            \label{fig:ca-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord1.png}
            \caption{}
            \label{fig:bon-reord1}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/bon-reord2.png}
            \caption{}
            \label{fig:bon-reord2}
    \end{subfigure}
    \caption{Southern Women Affiliation matrix with re-ordered rows and columns: (a) row and columns re-ordered according to the person and group score on the first CA dimension, (b) row and columns re-ordered according to person and group Eigenvector centrality score, (c) row and columns re-ordered according to person and group score on the second Bonacich dimension.}
    \label{fig:ca-v-bon}
\end{figure}

@@

@@

This suggests that the eigenvector centralities do encode a kind of similarity between pairs of people and pairs of groups; it just happens to be the unweighted similarity scores given by their raw number of shared memberships (for people) and shared members (for groups). In this respect, the dual centrality eigenvector score mixes an ordinal rank criterion with a similarity criterion since people with lots of memberships will end up being similar to lots of other people (and the same for large groups). Also, just like with CA and the degree-weighted projections, higher order axes of the Bonacich centrality eigendecomposition of the Breiger projection matrices should also encode forms of similarity between pairs of people and pairs of groups, independently of the main axis dominated by overall person activity and group size.

While plotting the Bonacich centrality against higher eigenvectors is not necessarily a substantively unmotivated practice \citep[see][for a discussion in the one-mode network case]{iacobucci2017eigenvector} the exact meaning of this mysterious second eigenvector of the Breiger projection matrices remains unclear, especially in the two-mode case. We can use the same ``a posteriori blockmodeling'' approach we used before to contrast CA and the Bonacich centrality analysis to shed light on this issue \citep{wasserman1990correspondence}. Figure~\ref{fig:bon-reord2} shows the affiliation matrix blocked according to the \textit{second} Bonacich dimension, and Figure~\ref{fig:bon-sim} (c) shows the ``Bonacich Correspondence Plot'' with the usual eigenvector centrality score on the x-axis plotted against the second eigenvector of the Breiger projection matrix for both persons and groups on the y-axis.

We can see from Figure~\ref{fig:bon-reord2} that the second Bonacich dimension reveals a \textit{dual triangular} structure, revealing two secondary core and peripheries that are not encoded in either the first Bonacich or CA dimensions. Interestingly, this partition is almost equivalent to that obtained in \citet[Table 5]{everett2013dual} by clustering a version of the projection matrices of the Southern Women data where the proximities were designed to detect \textit{structural equivalence} (e.g., the cell entry in each projection matrix was the proportion of matches between each pair of people and groups).\footnote{As \citet[208]{everett2013dual} define it, ``[p]roportion matching is the number of times two profiles have the same entry (either zeros or ones) divided by the length of the profile.''} In Everett and Borgatti's \citeyearpar[208]{everett2013dual} clustering solution, ``[t]he women are then split into (a) those that attend four or fewer events and consequently attend very few of the peripheral events (Pearl, Ruth, Verne, Myrna [sic], Dorothy, Olivia, Flora), and (b) those that attend core events and the first group of peripheral events (Evelyn, Laura, Theresa, Brenda, Charlotte, Francis, Eleanor), and (c) those that attend core events and the second group of peripheral events (Helen, Katherine, Sylvia, Nora).'' Note that the only disagreement between Everett and Borgatti's partition and the three-block solution suggested by Figure~\ref{fig:bon-reord2} is the position of Myrna/Myra's, who is assigned to Borgatti and Everett's third group in the a posteriori blocking obtained from Bonacich's second dimension. This may be partly due to the discrepant position of ``core event'' E9 (according to the main eigenvector dimension) which is grouped with the other peripheral events \{E10, E11, E12, E13, E14\} in Figure~\ref{fig:bon-reord2}.

This analysis points to a new way to interpret the second Bonacich dimension, namely, as a way to reveal secondary core-periphery partitions net of a main core-periphery partition in the two-mode network (if any exists). Furthermore, the similarities between persons and groups defined over the two-dimensional Bonacich space, shown in Figure~\ref{fig:bon-sim} (c), can thus be used to reveal structurally equivalent clusters of persons and events similar to those obtained by \citet[Table 5]{everett2013dual} using UCINET's combinatorial optimization routine specifying a three cluster solution of the (proportion matching) projection matrix for persons and groups. Figure~\ref{fig:bonclust} shows the results of a k-means analysis using the first two Bonacich dimensions to define between-person and groups distances, revealing a three cluster solution equivalent to that obtained by Borgatti and Everett, with the exception of Myrna/Myra's position as discussed earlier. This suggests that the Bonacich approach to similarity clustering is distinct from CA in being useful for revealing structural equivalence, thus making the Bonacich dimensions a more faithful basis for a low-rank reconstruction of the original ``input data'' than CA \citep{desposito2014comparison}.
